\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{hyperref}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

\newcommand{\inner}[2]{\langle #1, #2 \rangle}
\newcommand{\norm}[1]{\left\| #1 \right\|}
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\F}{\mathbb{F}}

\title{Inner Product Spaces}
\author{Math Foundations --- Node 08}
\date{}

\begin{document}
\maketitle

\section{Motivation}

In $\R^n$ the dot product $u \cdot v = \sum_i u_i v_i$ encodes both
length, via $\norm{u}^2 = u \cdot u$, and angle, via the relation
$\cos\theta = (u \cdot v)/(\norm{u}\,\norm{v})$. An \emph{inner product
space} is the abstraction that lets us carry these geometric notions
into any vector space, including spaces of functions, sequences, and
random variables. Linear algebra without an inner product gives us
linear maps; with one, it gives us geometry.

\section{Definitions}

Let $V$ be a vector space over $\F \in \{\R, \C\}$.

\begin{definition}[Inner product]
An \emph{inner product} on $V$ is a map
$\inner{\cdot}{\cdot} : V \times V \to \F$ satisfying, for all
$u, v, w \in V$ and $\alpha \in \F$:
\begin{enumerate}
  \item \textbf{Conjugate symmetry:} $\inner{u}{v} = \overline{\inner{v}{u}}$.
  \item \textbf{Linearity in the first argument:}
        $\inner{\alpha u + w}{v} = \alpha \inner{u}{v} + \inner{w}{v}$.
  \item \textbf{Positive definiteness:} $\inner{u}{u} \ge 0$, with
        equality iff $u = 0$.
\end{enumerate}
A vector space equipped with an inner product is called an
\emph{inner product space}.
\end{definition}

Note that over $\R$, conjugate symmetry collapses to symmetry, and
linearity in the first argument together with symmetry gives linearity
in the second argument as well. Over $\C$, the second argument is
\emph{conjugate}-linear; this is sometimes called the ``physicist's
convention'' when the conjugate sits in the first slot --- we follow
the mathematician's convention here.

\begin{definition}[Induced norm]
The \emph{induced norm} on an inner product space $(V, \inner{\cdot}{\cdot})$ is
$$\norm{v} := \sqrt{\inner{v}{v}}, \qquad v \in V.$$
Positive definiteness makes this well-defined and non-negative.
\end{definition}

\begin{definition}[Orthonormal set]
A finite or countable family $\{e_i\}_{i \in I} \subset V$ is called
\emph{orthonormal} if
$$\inner{e_i}{e_j} = \delta_{ij}
  = \begin{cases} 1 & i = j, \\ 0 & i \neq j.\end{cases}$$
\end{definition}

\section{The Cauchy-Schwarz inequality}

\begin{theorem}[Cauchy-Schwarz]
\label{thm:cs}
Let $(V, \inner{\cdot}{\cdot})$ be an inner product space over $\F$.
For all $u, v \in V$,
\begin{equation}
\label{eq:cs}
\bigl| \inner{u}{v} \bigr| \;\le\; \norm{u}\,\norm{v},
\end{equation}
with equality if and only if $u$ and $v$ are linearly dependent.
\end{theorem}

\begin{proof}
If $v = 0$ both sides of \eqref{eq:cs} are zero and there is nothing
to prove, so assume $v \neq 0$. The trick is to consider, for a scalar
$t \in \F$ to be chosen, the non-negative quantity
$$0 \le \norm{u - t v}^2 = \inner{u - t v}{u - t v}.$$
Expanding using linearity in the first slot and conjugate linearity in
the second,
\begin{align*}
\inner{u - tv}{u - tv}
  &= \inner{u}{u} - t\,\inner{v}{u} - \bar t\,\inner{u}{v} + |t|^2 \inner{v}{v} \\
  &= \norm{u}^2 - t\,\overline{\inner{u}{v}} - \bar t\,\inner{u}{v}
     + |t|^2 \norm{v}^2.
\end{align*}
Choose $t = \inner{u}{v} / \norm{v}^2$. Then $\bar t \inner{u}{v} =
|\inner{u}{v}|^2 / \norm{v}^2$ and $t\,\overline{\inner{u}{v}} =
|\inner{u}{v}|^2 / \norm{v}^2$, while $|t|^2 \norm{v}^2 =
|\inner{u}{v}|^2 / \norm{v}^2$. Substituting,
$$0 \le \norm{u}^2 - \frac{|\inner{u}{v}|^2}{\norm{v}^2}.$$
Multiplying through by $\norm{v}^2 > 0$ and rearranging,
$$|\inner{u}{v}|^2 \le \norm{u}^2 \norm{v}^2,$$
and taking square roots gives \eqref{eq:cs}.

For the equality case: equality forces $\norm{u - t v}^2 = 0$, so by
positive definiteness $u = tv$, i.e.\ $u$ and $v$ are linearly
dependent. Conversely if $u = \lambda v$ then both sides of
\eqref{eq:cs} equal $|\lambda|\,\norm{v}^2$.
\end{proof}

\begin{proposition}[Triangle inequality]
For all $u, v \in V$, $\norm{u + v} \le \norm{u} + \norm{v}$.
\end{proposition}

\begin{proof}
Expand $\norm{u+v}^2 = \norm{u}^2 + 2\,\Re\,\inner{u}{v} + \norm{v}^2$,
bound $\Re\,\inner{u}{v} \le |\inner{u}{v}| \le \norm{u}\,\norm{v}$ by
Theorem~\ref{thm:cs}, and take square roots.
\end{proof}

\section{Examples}

\begin{example}[Euclidean space]
$V = \R^n$, $\inner{u}{v} = \sum_{i=1}^n u_i v_i$. The induced norm is
the usual Euclidean length, and Cauchy-Schwarz is the classical
$\bigl(\sum u_i v_i\bigr)^2 \le \bigl(\sum u_i^2\bigr)\bigl(\sum v_i^2\bigr)$.
\end{example}

\begin{example}[$L^2$ on the unit interval]
$V = C[0,1]$ with $\inner{f}{g} = \int_0^1 f(x)\,g(x)\,dx$. Cauchy-Schwarz
becomes
$$\left|\int_0^1 f g\,dx\right|
   \le \left(\int_0^1 f^2\,dx\right)^{1/2}
       \left(\int_0^1 g^2\,dx\right)^{1/2},$$
a special case of H\"older's inequality with $p = q = 2$.
\end{example}

\section{Where this leads}

Cauchy-Schwarz is the inequality that lets us define $\cos\theta =
\inner{u}{v}/(\norm{u}\norm{v}) \in [-1, 1]$. The construction of
orthonormal bases via Gram-Schmidt, and orthogonal projection onto a
closed subspace, both rest on this single inequality. In ML, every
similarity score that takes the form of a dot product --- attention
logits, contrastive loss similarities, cosine similarity --- is a
direct corollary.

\end{document}
