\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}

\title{It\^o Calculus}
\author{math-foundations / concept 34}
\date{}

\begin{document}
\maketitle

\section{Setting}

Fix a filtered probability space $(\Omega, \mathcal{F}, (\mathcal{F}_t)_{t\geq 0}, \mathbb{P})$
satisfying the usual conditions and supporting a standard Brownian motion $W = (W_t)_{t\geq 0}$
adapted to $(\mathcal{F}_t)$. Sample paths of $W$ are almost surely continuous and nowhere
differentiable, with unbounded total variation on every compact interval.

\begin{definition}[Quadratic variation]
For a stochastic process $X = (X_t)$ and a partition $\Pi = \{0 = t_0 < t_1 < \cdots < t_n = t\}$
write $\|\Pi\| = \max_i (t_i - t_{i-1})$. The \emph{quadratic variation} of $X$ on $[0,t]$ is
\[
  [X]_t \;:=\; \lim_{\|\Pi\|\to 0} \sum_{i=1}^n \bigl(X_{t_i} - X_{t_{i-1}}\bigr)^2,
\]
provided the limit exists in probability. For Brownian motion, $[W]_t = t$ a.s.
\end{definition}

\begin{definition}[It\^o integral]
Let $H = (H_t)$ be a progressively measurable process with $\mathbb{E}\int_0^T H_s^2\,ds < \infty$.
The \emph{It\^o integral} of $H$ against $W$ is the $L^2$-limit
\[
  \int_0^t H_s\,dW_s \;:=\; \lim_{\|\Pi\|\to 0} \sum_{i=1}^n H_{t_{i-1}}\bigl(W_{t_i} - W_{t_{i-1}}\bigr).
\]
Evaluation at the \emph{left endpoint} $t_{i-1}$ is essential: it makes $t \mapsto \int_0^t H\,dW$
a continuous square-integrable martingale, with It\^o isometry
$\mathbb{E}\bigl[(\int_0^t H\,dW)^2\bigr] = \mathbb{E}\int_0^t H_s^2\,ds$.
\end{definition}

\begin{definition}[It\^o process]
A process $X$ is an \emph{It\^o process} if
\[
  X_t = X_0 + \int_0^t b_s\,ds + \int_0^t \sigma_s\,dW_s,
\]
for adapted $b, \sigma$ with $\int_0^T |b_s|\,ds < \infty$ and $\int_0^T \sigma_s^2\,ds < \infty$ a.s.
We write this symbolically as $dX_t = b_t\,dt + \sigma_t\,dW_t$.
\end{definition}

\section{It\^o's Lemma}

\begin{theorem}[It\^o's lemma, 1D]
\label{thm:ito}
Let $X_t = X_0 + \int_0^t b_s\,ds + \int_0^t \sigma_s\,dW_s$ be an It\^o process and let
$f \in C^2(\mathbb{R})$. Then $f(X_t)$ is again an It\^o process and
\[
  df(X_t) \;=\; f'(X_t)\,dX_t + \tfrac{1}{2}\,f''(X_t)\,\sigma_t^2\,dt.
\]
Equivalently, in integral form,
\[
  f(X_t) - f(X_0) \;=\; \int_0^t f'(X_s)\,b_s\,ds + \int_0^t f'(X_s)\,\sigma_s\,dW_s
  + \tfrac{1}{2}\int_0^t f''(X_s)\,\sigma_s^2\,ds.
\]
\end{theorem}

\begin{proof}[Proof sketch via second-order Taylor expansion and $(dW_t)^2 = dt$]
Fix a partition $\Pi = \{0 = t_0 < \cdots < t_n = t\}$ and write $\Delta_i X = X_{t_i} - X_{t_{i-1}}$,
$\Delta_i t = t_i - t_{i-1}$, $\Delta_i W = W_{t_i} - W_{t_{i-1}}$. By telescoping and a
second-order Taylor expansion of $f$ around $X_{t_{i-1}}$,
\begin{align*}
f(X_t) - f(X_0)
 &= \sum_{i=1}^n \Bigl[ f(X_{t_i}) - f(X_{t_{i-1}}) \Bigr] \\
 &= \sum_i f'(X_{t_{i-1}})\,\Delta_i X
   + \tfrac{1}{2}\sum_i f''(X_{t_{i-1}})\,(\Delta_i X)^2 + R_\Pi,
\end{align*}
where $R_\Pi$ collects the higher-order Taylor remainder, controlled by the modulus of continuity
of $f''$ on a compact set containing the path of $X$.

\smallskip\noindent\textbf{First-order term.}
Since $\Delta_i X = \int_{t_{i-1}}^{t_i} b_s\,ds + \int_{t_{i-1}}^{t_i} \sigma_s\,dW_s$,
the sum $\sum_i f'(X_{t_{i-1}})\,\Delta_i X$ converges in $L^2$ to
\[
\int_0^t f'(X_s)\,b_s\,ds + \int_0^t f'(X_s)\,\sigma_s\,dW_s,
\]
by definition of the It\^o integral and dominated convergence; the left-endpoint evaluation
$f'(X_{t_{i-1}})$ is exactly what the It\^o construction requires.

\smallskip\noindent\textbf{Second-order term --- the key identity $(dW_t)^2 = dt$.}
Expand
\[
(\Delta_i X)^2 = b_{t_{i-1}}^2 (\Delta_i t)^2 + 2 b_{t_{i-1}}\sigma_{t_{i-1}}(\Delta_i t)(\Delta_i W)
  + \sigma_{t_{i-1}}^2 (\Delta_i W)^2 + o(\Delta_i t)\text{-corrections}.
\]
The $(\Delta_i t)^2$ contributions sum to $O(\|\Pi\| \cdot t) \to 0$. The cross-terms
$(\Delta_i t)(\Delta_i W)$ have $L^2$-norm $O(\Delta_i t^{3/2})$, summing to $O(\|\Pi\|^{1/2}) \to 0$.
The surviving piece is
\[
  \sum_i f''(X_{t_{i-1}})\,\sigma_{t_{i-1}}^2\,(\Delta_i W)^2.
\]
Because $\mathbb{E}[(\Delta_i W)^2] = \Delta_i t$ and $\mathrm{Var}[(\Delta_i W)^2] = 2(\Delta_i t)^2$,
a direct $L^2$ computation shows that $(\Delta_i W)^2$ may be replaced by $\Delta_i t$ in the limit:
\[
  \sum_i f''(X_{t_{i-1}})\,\sigma_{t_{i-1}}^2 (\Delta_i W)^2
  \;\xrightarrow[\|\Pi\|\to 0]{L^2}\;
  \int_0^t f''(X_s)\,\sigma_s^2\,ds.
\]
This is precisely the heuristic identity $(dW_t)^2 = dt$ made rigorous, and it is an instance of
the quadratic variation $[W]_t = t$.

\smallskip\noindent\textbf{Remainder.}
$R_\Pi \to 0$ in probability because $f''$ is continuous and the path of $X$ is bounded on $[0,t]$
almost surely. Collecting the surviving pieces yields the formula stated.
\end{proof}

\begin{remark}[Why the left endpoint?]
The proof shows the limit of $\sum H_{t_{i-1}}\Delta_i W$ depends on the choice of evaluation
point. Replacing $H_{t_{i-1}}$ with the midpoint $H_{(t_{i-1}+t_i)/2}$ gives the
\emph{Stratonovich} integral, which obeys the classical chain rule
$df(X_t) = f'(X_t) \circ dX_t$ but is not a martingale. The right-endpoint choice peeks at the
future and produces a different, anticipating object. It\^o's left-endpoint choice gives a
martingale, at the cost of the second-order correction term in It\^o's lemma.
\end{remark}

\begin{remark}[Canonical example]
With $f(x) = x^2$ and $X_t = W_t$, It\^o's lemma yields
\[
d(W_t^2) = 2W_t\,dW_t + dt \quad\Longrightarrow\quad
\int_0^t W_s\,dW_s = \tfrac{1}{2}(W_t^2 - t).
\]
The classical guess $\tfrac{1}{2}W_t^2$ is off by exactly the It\^o correction $-t/2$.
\end{remark}

\end{document}
