\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{hyperref}

\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{lemma}{Lemma}

\title{Information Geometry: The Fisher Metric}
\author{math-foundations / concept 40}
\date{}

\begin{document}
\maketitle

\subsection*{First, an example}

Let $X \sim \mathcal{N}(\mu, 1)$ with unknown mean $\mu \in \mathbb{R}$ and unit variance.
The log-density is
\[
\ell(\mu) \;=\; -\tfrac{1}{2}(x - \mu)^2 \;-\; \tfrac{1}{2}\log(2\pi).
\]
\emph{Read aloud:} ell of mu equals minus one-half times x minus mu squared, minus one-half log two pi.
The score is $\partial_\mu \ell = x - \mu$, and $\mathbb{E}[x - \mu] = 0$ — the score has mean zero,
always. The Fisher information is
\[
F(\mu) \;=\; \mathbb{E}\!\left[(\partial_\mu \ell)^2\right] \;=\; \mathbb{E}[(x-\mu)^2] \;=\; 1.
\]
\emph{Read aloud:} F of mu equals expectation of partial-mu ell squared equals expectation of x minus mu squared equals one.
Now compare two nearby Gaussians: a direct calculation gives
\[
D_{\mathrm{KL}}\!\left(\mathcal{N}(0,1) \,\|\, \mathcal{N}(\epsilon, 1)\right) \;=\; \tfrac{1}{2}\,\epsilon^2,
\]
\emph{Read aloud:} K-L divergence from N zero one to N epsilon one equals one-half epsilon squared.
which is exactly $\tfrac{1}{2}\, F(\mu)\, \epsilon^2$. The Fisher metric \emph{is} the
second-order curvature of KL — that is the entire content of this lesson, in one example.

\section{Setting}

Let $\{p_\theta : \theta \in \Theta\}$ be a smoothly parameterized family of
probability densities on a measurable space $\mathcal{X}$, with
$\Theta \subseteq \mathbb{R}^d$ open. We assume sufficient regularity to
interchange differentiation in $\theta$ with integration in $x$, and that
$\log p_\theta(x)$ is twice continuously differentiable in $\theta$ for almost
every $x$. Define the \emph{score} $s_\theta(x) := \nabla_\theta \log p_\theta(x)$.

\begin{definition}[Fisher information matrix]
The Fisher information matrix at $\theta$ is the $d \times d$ matrix
\[
F(\theta)
  \;=\; \mathbb{E}_{X \sim p_\theta}\!\left[s_\theta(X)\, s_\theta(X)^\top\right]
  \;=\; -\,\mathbb{E}_{X \sim p_\theta}\!\left[\nabla_\theta^{\,2} \log p_\theta(X)\right].
\]
\emph{Read aloud:} F of theta equals the expectation under X drawn from p-theta of the score times its transpose, which equals minus the expectation of the Hessian of log p-theta.
The two expressions agree under the regularity assumptions above.
\end{definition}

\begin{definition}[Statistical manifold]
A \emph{statistical manifold} is a pair $(\Theta, F)$ where $\Theta$ is a
smooth manifold parameterizing $\{p_\theta\}$ and $F$ is the Fisher information
matrix viewed as a Riemannian metric tensor: for $u, v \in T_\theta \Theta$,
$\langle u, v \rangle_\theta := u^\top F(\theta) v$.
\end{definition}

\begin{definition}[Natural gradient]
Given a smooth loss $L : \Theta \to \mathbb{R}$, the \emph{natural gradient} is
\[
\widetilde{\nabla} L(\theta) \;=\; F(\theta)^{-1} \nabla L(\theta),
\]
\emph{Read aloud:} natural-gradient L of theta equals F of theta inverse times the ordinary gradient of L at theta.
defined wherever $F(\theta)$ is positive definite. It is the steepest-descent
direction under the Fisher metric.
\end{definition}

\section{The score has mean zero}

\begin{lemma}\label{lem:score-mean}
$\mathbb{E}_{X \sim p_\theta}[s_\theta(X)] = 0.$
\end{lemma}

\begin{proof}
Since $\int p_\theta(x)\, dx = 1$ for every $\theta$, differentiating in $\theta$
and exchanging derivative and integral,
\[
0 \;=\; \nabla_\theta \!\int p_\theta(x)\, dx
  \;=\; \int \nabla_\theta p_\theta(x)\, dx
  \;=\; \int \frac{\nabla_\theta p_\theta(x)}{p_\theta(x)}\, p_\theta(x)\, dx
  \;=\; \mathbb{E}_{p_\theta}[s_\theta(X)]. \qedhere
\]
\emph{Read aloud:} zero equals the gradient in theta of the integral of p-theta of x dx, equals the integral of the gradient of p-theta dx, equals the integral of grad-p over p times p dx, equals the expectation of the score under p-theta.
\end{proof}

\section{Fisher is the local KL metric}

\begin{theorem}\label{thm:fisher-kl}
The second-order Taylor expansion of
$\epsilon \mapsto D_{\mathrm{KL}}(p_\theta \,\|\, p_{\theta + \epsilon})$ near
$\epsilon = 0$ is
\[
D_{\mathrm{KL}}\!\left(p_\theta \,\|\, p_{\theta + \epsilon}\right)
   \;=\; \tfrac{1}{2}\, \epsilon^\top F(\theta)\, \epsilon
   \;+\; O\!\left(\|\epsilon\|^3\right).
\]
\emph{Read aloud:} K-L divergence from p-theta to p-theta-plus-epsilon equals one-half epsilon transpose F of theta epsilon, plus big-O of norm epsilon cubed.
\end{theorem}

\begin{proof}
Define $\Phi(\epsilon) := D_{\mathrm{KL}}(p_\theta \,\|\, p_{\theta + \epsilon})
  = \mathbb{E}_{X \sim p_\theta}\!\left[\log p_\theta(X) - \log p_{\theta+\epsilon}(X)\right].$
Set $\ell(\epsilon, x) := \log p_{\theta + \epsilon}(x)$, so
$\Phi(\epsilon) = \mathbb{E}_{p_\theta}[\ell(0, X) - \ell(\epsilon, X)]$.

Taylor-expand $\ell(\epsilon, x)$ in $\epsilon$ at $0$ to second order, with
remainder $R(\epsilon, x) = O(\|\epsilon\|^3)$ uniformly on compact sets under
our smoothness hypotheses:
\[
\ell(\epsilon, x)
  \;=\; \ell(0, x)
  + \epsilon^\top \nabla_\epsilon \ell(0,x)
  + \tfrac{1}{2}\, \epsilon^\top \nabla_\epsilon^{\,2} \ell(0, x)\, \epsilon
  + R(\epsilon, x).
\]
\emph{Read aloud:} ell of epsilon comma x equals ell at zero plus epsilon transpose times the gradient at zero plus one-half epsilon transpose Hessian at zero epsilon, plus a remainder R of epsilon x.
Note $\nabla_\epsilon \ell(0,x) = s_\theta(x)$ and
$\nabla_\epsilon^{\,2} \ell(0,x) = \nabla_\theta^{\,2} \log p_\theta(x)$.
Substituting,
\[
\ell(0,x) - \ell(\epsilon, x)
  \;=\; -\,\epsilon^\top s_\theta(x)
  - \tfrac{1}{2}\, \epsilon^\top \nabla_\theta^{\,2} \log p_\theta(x)\, \epsilon
  - R(\epsilon, x).
\]
\emph{Read aloud:} ell at zero minus ell at epsilon equals minus epsilon transpose score, minus one-half epsilon transpose Hessian of log p-theta epsilon, minus the remainder.
Take expectation under $p_\theta$. By Lemma~\ref{lem:score-mean},
$\mathbb{E}_{p_\theta}[\epsilon^\top s_\theta(X)] = \epsilon^\top \mathbb{E}_{p_\theta}[s_\theta(X)] = 0$,
so the linear term vanishes. By the Hessian form of Fisher,
$-\mathbb{E}_{p_\theta}[\nabla_\theta^{\,2} \log p_\theta(X)] = F(\theta)$, so
\[
\Phi(\epsilon)
  \;=\; 0 \;+\; \tfrac{1}{2}\, \epsilon^\top F(\theta)\, \epsilon
  \;-\; \mathbb{E}_{p_\theta}[R(\epsilon, X)]
  \;=\; \tfrac{1}{2}\, \epsilon^\top F(\theta)\, \epsilon \;+\; O\!\left(\|\epsilon\|^3\right). \qedhere
\]
\emph{Read aloud:} Phi of epsilon equals zero plus one-half epsilon transpose F of theta epsilon minus the expectation of the remainder, equals one-half epsilon transpose F of theta epsilon plus big-O of norm epsilon cubed.
\end{proof}

\paragraph{Consequence.} The Hessian of $\epsilon \mapsto
D_{\mathrm{KL}}(p_\theta \| p_{\theta+\epsilon})$ at $\epsilon = 0$ is
$F(\theta)$. The Fisher information is therefore the unique tensor that
captures the local quadratic behaviour of KL divergence — justifying its role
as the natural Riemannian metric on the statistical manifold and as the
preconditioner in natural-gradient descent.

\end{document}
