\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{hyperref}

\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}

\DeclareMathOperator{\KL}{D_{\mathrm{KL}}}
\DeclareMathOperator{\E}{\mathbb{E}}

\title{Mutual Information}
\author{math-foundations / concept 39}
\date{}

\begin{document}
\maketitle

\subsection*{First, an example}

Two coin flips $X, Y \in \{0, 1\}$, each fair. Work in bits ($\log_2$).
\begin{itemize}
  \item \textbf{Independent.} If the coins are independent, knowing $X$ tells you
    nothing about $Y$, and $I(X; Y) = 0$.
  \item \textbf{Perfectly correlated.} If $Y = X$ always, then learning $X$ pins
    down $Y$ exactly: $I(X; Y) = H(X) = 1$ bit.
  \item \textbf{Noisy channel.} Suppose $P(Y = X) = 0.9$, $P(Y \ne X) = 0.1$. By
    symmetry $Y$ is still fair, so $H(Y) = 1$. The conditional entropy is the
    binary entropy of the flip probability,
    $H(Y \mid X) = H_b(0.1) \approx 0.469$. Hence
    $I(X; Y) = H(Y) - H(Y \mid X) \approx 1 - 0.469 = 0.531$ bits.
\end{itemize}
Mutual information interpolates smoothly between zero (independence) and the full
entropy of $X$ (functional dependence).

\section{Setting}

Let $X$ and $Y$ be random variables on a common probability space, taking values in
finite (or countable) alphabets $\mathcal{X}$ and $\mathcal{Y}$. Write $p(x, y)$ for
the joint probability mass function and $p(x), p(y)$ for the marginals. All
logarithms are natural unless otherwise stated; replace $\log$ with $\log_2$ to
work in bits.

\section{Definitions}

\begin{definition}[Mutual information, KL form]
The \emph{mutual information} of $X$ and $Y$ is
\[
  I(X; Y) \;:=\; \KL\!\bigl(p(x, y) \,\Vert\, p(x)\, p(y)\bigr)
              \;=\; \sum_{x \in \mathcal{X}} \sum_{y \in \mathcal{Y}}
                    p(x, y) \log \frac{p(x, y)}{p(x)\, p(y)},
\]
\emph{Read aloud:} ``the mutual information of $X$ and $Y$ is defined as the KL
divergence from the joint distribution $p(x, y)$ to the product of marginals
$p(x) p(y)$, which equals the double sum over $x$ in $\mathcal{X}$ and $y$ in
$\mathcal{Y}$ of $p$ of $x, y$ times log of $p$ of $x, y$ over $p$ of $x$ times
$p$ of $y$.''
with the conventions $0 \log 0 = 0$ and $0 \log (0/0) = 0$.
\end{definition}

\begin{definition}[Entropic form]
Equivalently,
\[
  I(X; Y) \;=\; H(X) - H(X \mid Y) \;=\; H(Y) - H(Y \mid X) \;=\; H(X) + H(Y) - H(X, Y),
\]
\emph{Read aloud:} ``mutual information of $X$ and $Y$ equals entropy of $X$
minus conditional entropy of $X$ given $Y$, which equals entropy of $Y$ minus
conditional entropy of $Y$ given $X$, which equals entropy of $X$ plus entropy
of $Y$ minus joint entropy of $X$ and $Y$.''
where $H(X) = -\sum_x p(x) \log p(x)$ and
$H(X \mid Y) = -\sum_{x, y} p(x, y) \log p(x \mid y)$.
\end{definition}

\section{Main theorem}

\begin{theorem}[Nonnegativity of mutual information]
\label{thm:mi-nonneg}
For all random variables $X, Y$,
\[
  I(X; Y) \;\ge\; 0,
\]
\emph{Read aloud:} ``the mutual information of $X$ and $Y$ is greater than or
equal to zero.''
with equality if and only if $X$ and $Y$ are independent.
\end{theorem}

\begin{proof}
By definition, $I(X; Y) = \KL(p_{XY} \,\Vert\, p_X p_Y)$, where $p_{XY}(x, y) = p(x, y)$
and $(p_X p_Y)(x, y) = p(x) p(y)$. Both functions are probability mass
functions on $\mathcal{X} \times \mathcal{Y}$: the first by construction, and
the second because
$\sum_{x, y} p(x) p(y) = \bigl(\sum_x p(x)\bigr)\bigl(\sum_y p(y)\bigr) = 1$.

From the KL-divergence node (concept 38, Gibbs' inequality), for any two pmfs
$p, q$ on a common alphabet $\mathcal{Z}$,
\begin{equation}
  \KL(p \,\Vert\, q) \;=\; \sum_z p(z) \log \frac{p(z)}{q(z)} \;\ge\; 0, \label{eq:gibbs}
\end{equation}
\emph{Read aloud:} ``the KL divergence from $p$ to $q$ equals the sum over $z$
of $p$ of $z$ times log of $p$ of $z$ over $q$ of $z$, and this is greater than
or equal to zero.''
with equality if and only if $p(z) = q(z)$ for every $z$ with $p(z) > 0$ (equivalently,
$p = q$ as measures). Applying~\eqref{eq:gibbs} with $\mathcal{Z} = \mathcal{X} \times \mathcal{Y}$,
$p = p_{XY}$, and $q = p_X p_Y$ gives $I(X; Y) \ge 0$ at once.

For the equality condition: by Gibbs, $I(X; Y) = 0$ iff $p(x, y) = p(x) p(y)$ for
every pair $(x, y) \in \mathcal{X} \times \mathcal{Y}$ — which is precisely
the definition of independence of $X$ and $Y$ (concept 21).
\end{proof}

\begin{corollary}[Symmetry]
$I(X; Y) = I(Y; X)$.
\end{corollary}

\begin{proof}
The KL form is manifestly symmetric: swapping the summation indices leaves both
$p(x, y)$ and $p(x) p(y)$ invariant. Equivalently, $H(X) + H(Y) - H(X, Y)$
is symmetric in $X$ and $Y$.
\end{proof}

\section{Worked example: binary symmetric channel}

Let $X \sim \mathrm{Bernoulli}(1/2)$ and $Y = X \oplus N$, with
$N \sim \mathrm{Bernoulli}(\varepsilon)$ independent of $X$, $\varepsilon \in [0, 1]$.
Then $Y \sim \mathrm{Bernoulli}(1/2)$, so $H(X) = H(Y) = \log 2$, and
$H(Y \mid X) = H_2(\varepsilon) := -\varepsilon \log \varepsilon - (1 - \varepsilon) \log(1 - \varepsilon)$.
Therefore
\[
  I(X; Y) \;=\; \log 2 - H_2(\varepsilon).
\]
\emph{Read aloud:} ``the mutual information of $X$ and $Y$ equals log 2 minus
the binary entropy of $\varepsilon$.''
At $\varepsilon \in \{0, 1\}$ the channel is deterministic and $I = \log 2$;
at $\varepsilon = 1/2$ the channel is pure noise and $I = 0$ — consistent with
Theorem~\ref{thm:mi-nonneg}, since $\varepsilon = 1/2$ makes $X \perp Y$.

\end{document}
