\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{hyperref}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}

\title{KL Divergence}
\author{Math Foundations --- Concept 38}
\date{}

\begin{document}
\maketitle

\section{Definition}

Let $p$ and $q$ be probability distributions on a common measurable space $(\mathcal{X}, \mathcal{F})$, with $p$ absolutely continuous with respect to $q$ (written $p \ll q$). The \emph{Kullback--Leibler divergence} of $p$ from $q$ is
\[
D_{\mathrm{KL}}(p \,\|\, q) \;=\; \mathbb{E}_{p}\!\left[\log \frac{dp}{dq}\right].
\]

\begin{definition}[Discrete KL divergence]
For probability mass functions $p, q$ on a countable set $\mathcal{X}$,
\[
D_{\mathrm{KL}}(p \,\|\, q) \;=\; \sum_{x \in \mathcal{X}} p(x) \log \frac{p(x)}{q(x)},
\]
with conventions $0 \log 0 = 0$ and $0 \log (0/0) = 0$. The sum is $+\infty$ if there exists $x$ with $p(x) > 0$ and $q(x) = 0$.
\end{definition}

\begin{definition}[Continuous KL divergence]
For probability densities $f_p, f_q$ on $\mathbb{R}^d$ (with respect to Lebesgue measure),
\[
D_{\mathrm{KL}}(p \,\|\, q) \;=\; \int_{\mathbb{R}^d} f_p(x) \log \frac{f_p(x)}{f_q(x)} \, dx,
\]
defined whenever the integrand is integrable and $\{f_q = 0\} \subseteq \{f_p = 0\}$ up to null sets.
\end{definition}

\begin{remark}
$D_{\mathrm{KL}}$ is \emph{not} symmetric and does \emph{not} satisfy the triangle inequality, so it is not a metric. Nonetheless it is non-negative, vanishes only when $p = q$, and behaves additively under independence, justifying its role as a ``relative information'' measure.
\end{remark}

\section{Gibbs' inequality}

\begin{theorem}[Non-negativity of KL]
\label{thm:gibbs}
For any pair of probability distributions $p, q$ on the same space with $p \ll q$,
\[
D_{\mathrm{KL}}(p \,\|\, q) \;\ge\; 0,
\]
with equality if and only if $p = q$ almost everywhere.
\end{theorem}

\begin{proof}
Write $r(x) = q(x) / p(x)$ on the support of $p$. By definition,
\[
D_{\mathrm{KL}}(p \,\|\, q) \;=\; \mathbb{E}_{p}\!\left[\log \frac{p(X)}{q(X)}\right] \;=\; \mathbb{E}_{p}\!\left[-\log r(X)\right].
\]
The function $\varphi(t) = -\log t$ is strictly convex on $(0, \infty)$. By Jensen's inequality applied to $\varphi$ and the random variable $r(X)$ under $p$,
\[
\mathbb{E}_{p}\!\left[-\log r(X)\right] \;\ge\; -\log \mathbb{E}_{p}\!\left[r(X)\right].
\]
We compute the right-hand side:
\[
\mathbb{E}_{p}\!\left[r(X)\right] \;=\; \int r(x) \, p(x) \, dx \;=\; \int \frac{q(x)}{p(x)} \, p(x) \, dx \;=\; \int q(x) \, dx \;=\; 1,
\]
since $q$ is a probability distribution. Therefore
\[
D_{\mathrm{KL}}(p \,\|\, q) \;\ge\; -\log 1 \;=\; 0.
\]
Because $\varphi$ is \emph{strictly} convex, Jensen's inequality is an equality iff $r(X)$ is $p$-almost-surely constant. Combined with $\mathbb{E}_p[r(X)] = 1$, this forces $r \equiv 1$ on the support of $p$, i.e.\ $p = q$ a.e. The discrete case is identical with integrals replaced by sums.
\end{proof}

\begin{remark}
Run with $\varphi = -\log$ replaced by an arbitrary convex $\varphi$ with $\varphi(1) = 0$, the same proof yields the family of \emph{$f$-divergences} (Csisz\'ar). KL corresponds to $\varphi(t) = t \log t$ (after a sign convention); other choices recover total variation, Hellinger, and $\chi^2$ divergences.
\end{remark}

\section{Consequences}

\begin{lemma}[Cross-entropy decomposition]
For all $p, q$ with $p \ll q$,
\[
H(p, q) \;=\; H(p) + D_{\mathrm{KL}}(p \,\|\, q),
\]
where $H(p) = -\mathbb{E}_p[\log p]$ and $H(p, q) = -\mathbb{E}_p[\log q]$.
\end{lemma}

By Theorem~\ref{thm:gibbs}, $H(p, q) \ge H(p)$ with equality iff $q = p$. Hence minimizing cross-entropy in $q$ for fixed $p$ is the same as minimizing KL divergence --- the foundational identity behind maximum likelihood estimation and the categorical cross-entropy loss.

\end{document}
